Make a README

### 2025-02-09 choose_license.qmd
Rework adding data set (again)
This reverses commit 6d3ca2b.
Instead of pointblank, which
has many dependencies, datawizard
and tinytable are used (again).
tinylabels is not necessary,
however, as adding labels can be
already achieved with datawizard
alone.

Main motivation was that basic
summary statistics should be
part of every data dictionary,
and althugh this can be achieved
with pointblank, it requires more
manual work. Also, value labels
are catered for automatically,
while with pointblank one would
first have to think where exactly
to indicate them (probably as
subsections to the "colu

For example, if you previously created the file `create_data_dictionary.R`,
you can now add the following comment to the beginning of the file,
replacing `<YEAR>` and `<YOUR NAME>` with the current year and your name
-- of course, you can also choose a different license:

```{.r filename="create_data_dictionary.R"}
# SPDX-FileCopyrightText: <YEAR> <YOUR NAME>
#
# SPDX-License-Identifier: CC0-1.0
```

You need to use `#` to start the comment
because this is the symbol that starts comment lines in R scripts.
Alternatively, you can use the [reuse tool](https://github.com/fsfe/reuse-tool)
to add these information for you.
After installing it with...

...you can add the copyright information using the following command
-- the current year will be added automatically:

```{.bash .code-overflow-wrap filename="Terminal"}
reuse annotate --copyright="<YOUR NAME>" --license="CC0-1.0" create_data_dictionary.R
```

In many cases, the reuse tool will figure out the appropriate comment style for you.
If this is not the case, as currently with Quarto files,
you can tell it directly which comment style to use (`html` in this case):

This adds the following header to `Manuscript.qmd`:

@@

Note that `Manuscript.qmd` was provided to you under [CC0\ 1.0][cc0],
which is what you indicate with the previous comment.
As you edited the file throughout this tutorial,
you may also add yourself.^[For licenses that require that modifications are indicated,
this is an easy way to comply with them.
Also, you do not need to provide your real name.]

@@

The manuscript and the bibliography file you downloaded for this tutorial
are made available to you under [CC0\ 1.0][cc0]
as is all code in this tutorial (e.g., for creating the data dictionary).
Therefore, you can apply any license to your modified version.
Please apply the license of your choice now,
using any of the methods mentioned above (except the `README`, which we will cover later).

@@

The bibliography (stored in "Bibliography.bib"), data dictionary (stored in "data_dictionary.html"), and the code used to create it (stored in "create_data_dictionary.R") by Florian Kohrt and <YOUR NAME> are available under CC0 1.0 <https://creativecommons.org/publicdomain/zero/1.0/>.
```

>>>>

For example, the file `Manuscript.qmd` was provided to you under [CC0\ 1.0][cc0].
You can indicate that by adding the following comment to the beginning of the file:

@@

You need to use `<!--` and `-->` to start and end the comment
because this is the convention that starts comment lines in Quarto files.
Alternatively, you can use the [reuse tool](https://github.com/fsfe/reuse-tool)
to add these information for you.
After installing it with...

...you can add the copyright information using the following command
-- the current year will be added automatically.
In many cases, the reuse tool will figure out the appropriate comment style for you.
If this is not the case, as currently with Quarto files,
you can tell it directly which comment style to use (`html` in this case):

As you edited the file throughout this tutorial,
you may also add yourself.^[For licenses that require that modifications are indicated,
this is an easy way to comply with them.
Also, you do not need to provide your real name.]

```{.bash .code-overflow-wrap filename="Terminal"}
reuse annotate --copyright="<YOUR NAME>" --license="CC0-1.0" --style=html Manuscript.qmd
```

@@

The manuscript and the bibliography file you downloaded for this tutorial
are made available to you under [CC0\ 1.0][cc0]
as is all code in this tutorial (e.g., for the data analysis).
Therefore, you can apply any license to your modified version.
Please apply the license of your choice now,
using any of the methods mentioned above (except the `README`, which we will cover later).

@@

The bibliography (stored in "Bibliography.bib") by Florian Kohrt and <YOUR NAME> is available under CC0 1.0 <https://creativecommons.org/publicdomain/zero/1.0/>.
```

### 2025-02-09 data.qmd
Rework adding data set (again)
This reverses commit 6d3ca2b.
Instead of pointblank, which
has many dependencies, datawizard
and tinytable are used (again).
tinylabels is not necessary,
however, as adding labels can be
already achieved with datawizard
alone.

Main motivation was that basic
summary statistics should be
part of every data dictionary,
and althugh this can be achieved
with pointblank, it requires more
manual work. Also, value labels
are catered for automatically,
while with pointblank one would
first have to think where exactly
to indicate them (probably as
subsections to the "colu

When distributing a data set, it is important to document the meaning (e.g., units)
and valid values of its variables.
This is typically done with a _data dictionary_ (also called a _codebook_).
In the following, we will demonstrate how to create a simple data dictionary
using the R package [`pointblank`][pointblank].
You can install it now using:

[pointblank]: https://rstudio.github.io/pointblank/

::: {.column-margin}
![pointblank: Data Validation and Organization of Metadata for Local and Remote Tables](images/pointblank.png){width=250px}
:::

```{.r filename="Console"}
renv::install("pointblank")
```

You can put the code that follows for creating the data dictionary
into a new file called `create_data_dictionary.R`.

First, we write down everything we know about the data set.
This includes:

- a general description of the data set
- descriptions of all columns
- valid values, where applicable

```{r, filename="create_data_dictionary.R"}
table_info <- c(
  title = "palmerpenguins::penguins",
  description = "Size measurements for adult foraging penguins near Palmer Station, Antarctica"
)
descriptions <- c(
  species = "a character string denoting penguin species",
  island = "a character string denoting island in Palmer Archipelago, Antarctica",
  bill_length_mm = "a number denoting bill length (millimeters)",
  bill_depth_mm = "a number denoting bill depth (millimeters)",
  flipper_length_mm = "an integer denoting flipper length (millimeters)",
  body_mass_g = "an integer denoting body mass (grams)",
  sex = "a character string denoting penguin sex",
  year = "an integer denoting the study year"
)

vals <- list(
  species = c("Adelie", "Gentoo", "Chinstrap"),
  island = c("Torgersen", "Biscoe", "Dream"),
  sex = c("male", "female"),
  year = c(2007, 2008, 2009)
)
```

@@

As the data set is from the R package `palmerpenguins`,
if you have it installed you can use the function `citation()`
to create such a reference:

```{r}
dat_source <- citation("palmerpenguins", auto = TRUE) |>
  format(bibtex = FALSE, style = "text")
```

Without the package `palmerpenguins` installed,
you can find a [suggested citation on its website][palmerpenguins-citation]
and run the following:

@@

```{r}
#| echo: false
#| class-output: "r code-overflow-wrap"
#| attr-output: 'filename="create_data_dictionary.R"'

# We have provided the data set as CSV file to the readers.
# Therefore, we cannot assume or require that readers have
# the R package palmerpenguins installed. Instead, we create
# the citation on our end and hide how we obtained it.
dat_source |>
  gsub("\\n", " ", x = _) |>
  gsub("\"", "\\\\\"", x = _) |> # Escape any quotation marks
  paste0("dat_source <- \"", ... = _, "\"") |>
  cat()
```

Then, we use `pointblank` to create a data dictionary with this information.

```{r, filename="create_data_dictionary.R"}
vals <- sapply(vals, \(x) {
  paste0(
    "(",
    knitr::combine_words(x, and = " or ", before = "`", after = "`"),
    ")"
  )
})

dict <- pointblank::create_informant(
  dat,
  tbl_name = NA,
  label = table_info[["title"]],
  lang = "en"
) |>
  pointblank::info_tabular(
    Description = table_info[["description"]],
    Source = dat_source
  ) |>
  pointblank::info_columns_from_tbl(stack(descriptions)[2:1]) |>
  pointblank::info_columns_from_tbl(stack(vals)[2:1]) |>
  pointblank::get_informant_report(
    title = "Data Dictionary for `data.csv`"
  )

dict
```

Finally, we can store the data dictionary inside an HTML file
and put the HTML file into the project folder as well.

```{.r filename="create_data_dictionary.R"}
pointblank::export_report(dict, filename = "data_dictionary.html")
```

For a more elaborate introduction to `pointblank`,
you can read its "[Intro to Information Management][pointblank-intro]".

[pointblank-intro]: https://rstudio.github.io/pointblank/articles/INFO-1.html

@@

This example demonstrates how the title and description of the data set,
the description of the variables and their valid values are stored in a machine-readable way.
As before, we also provide a reference to the source.

@@

@@

```{.r filename="Console"}
# Read data and create schema
dat_filename <- "data.csv"
dat <- read.csv(dat_filename)
dat_schema <- frictionless::create_schema(dat)

# Add descriptions to the fields
dat_schema$fields <- lapply(dat_schema$fields, \(x) {
  c(x, description = descriptions[[x$name]])
})

# Record valid values
dat_schema$fields <- lapply(dat_schema$fields, \(x) {
  if (x$name %in% names(vals)) {
    modifyList(x, list(constraints = list(enum = vals[[x$name]])))
  } else {
    x
  }
})

>>>>

Whether or not distributing the data set,
it is important to document the meaning (e.g., units)
and values of its variables.
This is typically done with a _data dictionary_ (also called a _codebook_).
In the following, we will demonstrate how to create a simple data dictionary
using the R packages [`datawizard`][datawizard] and [`tinytable`][tinytable].
You can install them now using:

[datawizard]: https://easystats.github.io/datawizard/
[tinytable]: https://vincentarelbundock.github.io/tinytable/

```{.r filename="Console"}
renv::install(c(
  "datawizard",
  "tinytable"
))
```

You can put the code that follows in a separate document.
Create it by clicking on _File_ > _New File_  > _Quarto Document..._.
Choose a title such as `Data Dictionary`,
select _HTML_ as format,
uncheck the use of the visual markdown editor, and click on _Create_.
Remove everything except the YAML header (between the `---`).
To make the HTML file self-contained,
also set `embed-resources: true` such that the YAML header looks as follows:

```{.yml filename="data_dictionary.qmd"}
---
title: "Data Dictionary"
format:
  html:
    embed-resources: true
---
```

Then, save it as `data_dictionary.qmd` by clicking on _File_ > _Save_.

To create the actual data dictionary, first write a description for all columns
so others can understand what the variable names mean.
Where necessary, also document their value
-- this is especially important if their meaning is non-obvious.
In the following, we demonstrate this by storing the penguins' binomial name
along with the English name.

``````{cat}
#| engine.opts = list(file = "_data_dictionary.qmd"),
#| class.source = "md",
#| filename = "data_dictionary.qmd"

```{r}
#| echo: false

vars <- c(
  species = "a character string denoting penguin species",
  island = "a character string denoting island in Palmer Archipelago, Antarctica",
  bill_length_mm = "a number denoting bill length (millimeters)",
  bill_depth_mm = "a number denoting bill depth (millimeters)",
  flipper_length_mm = "an integer denoting flipper length (millimeters)",
  body_mass_g = "an integer denoting body mass (grams)",
  sex = "a character string denoting penguin sex",
  year = "an integer denoting the study year"
)

vals <- list(
  species = c(
    Adelie = "Pygoscelis adeliae",
    Gentoo = "Pygoscelis papua",
    Chinstrap = "Pygoscelis antarcticus"
  )
)
```
``````

Then, load the data and use `datawizard`
to add the descriptions to the `data.frame`:[^not-permanent]

[^not-permanent]: Note that the code provided does not alter the data file
-- no description will be added to `data.csv`.
The descriptions are only added to a (temporary) copy of the data set within R
to create the data dictionary.

::: {.column-margin}
![datawizard: Easy Data Wrangling and Statistical Transformations](images/datawizard.png){width=250px}
:::

``````{cat}
#| engine.opts = list(file = "_data_dictionary.qmd", append = TRUE),
#| class.source = "md",
#| filename = "data_dictionary.qmd"

```{r}
#| echo: false

for (x in names(vars)) {
  if (x %in% names(vals)) {
    dat <- datawizard::assign_labels(
      dat,
      select = I(x),
      variable = vars[[x]],
      values = vals[[x]]
    )
  } else {
    dat <- datawizard::assign_labels(
      dat,
      select = I(x),
      variable = vars[[x]]
    )
  }
}
```

``````

Next, you can use `datawizard` again to create
the data dictionary containing the descriptions,
but also some other information about each variable
(e.g., the number of missing values).
Finally, you can use `tinytable` to print the table.

::: {.column-margin}
![tinytable: Simple and Configurable Tables](images/tinytable.svg){width=250px}
:::

``````{cat}
#| engine.opts = list(file = "_data_dictionary.qmd", append = TRUE),
#| class.source = "md",
#| filename = "data_dictionary.qmd"

```{r}
#| echo: false
#| column: "body-outset"

(dict <- datawizard::data_codebook(dat) |>
  subset(select = -c(ID, .row_id)) |>
  tinytable::tt() |>
  tinytable::format_tt(replace = "")) # hide NAs
```

``````

```{r}
#| child: "_data_dictionary.qmd"

```

@@

The data set is from the R package `palmerpenguins`.
If you had it installed
you could use the function `citation()` to create such a reference:

```{r}
#| label: "data-citation"
#| output: false

citation("palmerpenguins", auto = TRUE) |>
  format(bibtex = FALSE, style = "text")
```

Without the package `palmerpenguins` installed,
you can find a [suggested citation on its website][palmerpenguins-citation]
and add that to your data dictionary:

@@

```{r}
#| ref.label = "data-citation",
#| render = function(x, options) gsub("\\n", " ", x = x),
#| echo = FALSE,
#| class.output = "md code-overflow-wrap",
#| attr.output = 'filename="data_dictionary.qmd"'

# This chunk takes the output from the chunk "data-citation"
# and renders it with all newlines replaced by whitespaces.
```

Finally, you can render the data dictionary by running the following:

```{.bash filename="Terminal"}
quarto render data_dictionary.qmd
```

This should create the file `data_dictionary.html`
which you open and view in your web browser.

@@

This example demonstrates how the title and description of the data set,
the description of the variables and their valid values are stored in a machine-readable way.
We'll reuse the descriptions we already created[^value-labels] and add a few others.

[^value-labels]: Unfortunately, the descriptions of values are not reused in this example,
as they are [not supported][enum-labels] by the specification we are using.

[enum-labels]: https://specs.frictionlessdata.io/patterns/#table-schema-enum-labels-and-ordering

First, store the title and description of the data set as a whole:

```{.r filename="Console"}
table_info <- c(
  title = "penguins data set",
  description = "Size measurements for adult foraging penguins near Palmer Station, Antarctica"
)
```

As before, also provide a reference to the source.

@@

Next, create a list of the categorical variables' valid values:

```{.r filename="Console"}
valid_vals <- list(
  species = c("Adelie", "Gentoo", "Chinstrap"),
  island = c("Torgersen", "Biscoe", "Dream"),
  sex = c("male", "female"),
  year = c(2007, 2008, 2009)
)
```

Finally, store the descriptions of the variables we already created earlier:

```{.r filename="Console"}
vars <- c(
  species = "a character string denoting penguin species",
  island = "a character string denoting island in Palmer Archipelago, Antarctica",
  bill_length_mm = "a number denoting bill length (millimeters)",
  bill_depth_mm = "a number denoting bill depth (millimeters)",
  flipper_length_mm = "an integer denoting flipper length (millimeters)",
  body_mass_g = "an integer denoting body mass (grams)",
  sex = "a character string denoting penguin sex",
  year = "an integer denoting the study year"
)
```

@@

```{.r filename="Console"}
# Install {frictionless} and the required dependency {stringi}
renv::install(c(
  "frictionless",
  "stringi"
))

# Read data and create schema
dat_filename <- "data.csv"
dat <- read.csv(dat_filename)
dat_schema <- frictionless::create_schema(dat)

# Add descriptions to the fields
dat_schema$fields <- lapply(dat_schema$fields, \(x) {
  c(x, description = vars[[x$name]])
})

# Record valid values
dat_schema$fields <- lapply(dat_schema$fields, \(x) {
  if (x[["name"]] %in% names(valid_vals)) {
    modifyList(x, list(constraints = list(enum = valid_vals[[x$name]])))
  } else {
    x
  }
})

### 2025-02-09 make_readme.qmd
Rework adding data set (again)
This reverses commit 6d3ca2b.
Instead of pointblank, which
has many dependencies, datawizard
and tinytable are used (again).
tinylabels is not necessary,
however, as adding labels can be
already achieved with datawizard
alone.

Main motivation was that basic
summary statistics should be
part of every data dictionary,
and althugh this can be achieved
with pointblank, it requires more
manual work. Also, value labels
are catered for automatically,
while with pointblank one would
first have to think where exactly
to indicate them (probably as
subsections to the "colu

```txt
Finding R package dependencies ... Done!
── Install scripts ──────────────────────── Fedora 40 ── 
dnf install -y make pandoc libcurl-devel \
  openssl-devel libicu-devel v8-devel libxml2-devel \
  git

── Packages and their system dependencies ──────────────
blastula  – pandoc
curl      – libcurl-devel, openssl-devel
fs        – make
knitr     – pandoc
openssl   – openssl-devel
remotes   – git
rmarkdown – pandoc
sass      – make
stringi   – libicu-devel
V8        – v8-devel
xml2      – libxml2-devel
```

@@

The other system dependencies are libraries,
and the method to identify the installed version varies between operating systems.
For example, on Fedora Linux, one can run the following:

```{.bash .code-overflow-wrap filename="Terminal"}
rpm -q libcurl-devel openssl-devel libicu-devel v8-devel libxml2-devel
```

However, they are only relevant if the R packages were installed from source,
rather than as binary package.

@@

- `Bibliography.bib`: bibliographic references used in the manuscript
- `data.csv`: a data set containing the simplified `palmerpenguins` data
- `data_dictionary.html`: a dictionary to the data file,
  created using `create_data_dictionary.R`

>>>>

```txt
── Install scripts ────────────────── Fedora 40 ── 
dnf install -y make pandoc git

── Packages and their system dependencies ────────
fs        – make
knitr     – pandoc
remotes   – git
rmarkdown – pandoc
sass      – make
```

@@

However, this does not work for all system dependencies.
Specifically, it does not work for libraries -- software that is not supposed to be run on its own.
Identifying their version is beyond the scope of this tutorial.

@@

- `Bibliography.bib`: bibliographic references used in the manuscript
- `data.csv`: a data set containing the simplified `palmerpenguins` data
- `data_dictionary.html`: a dictionary to the data file,
  created using `data_dictionary.qmd`
